A novel framework for linking dynamic experiences to memories reveals event-like structure in naturalistic episodic recall

### 2019-07-22 paper/main.tex
saving progress on list learning discussion section

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events using HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented'' events from the original video, and a sequence of (potentially differently ordered) ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{WelcBurn24, PostPhil65, AtkiShif68}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for 2 of these analyses (probability of first recall and lag-conditional response probability curves) we observe patterns comparable to classic effects from the list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}C). In contrast, we do not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather we see higher memory for specific events distributed somewhat evenly throughout the video.

@@

Our work casts remembering as reproducing (behaviorally and neurally) the topic trajectory, or shape, of the original experience.  This view draws inspiration from prior work aimed at elucidating the neural and behavioral underpinnings of how we process dynamic naturalistic experiences and remember them later.  One approach to identifying neural responses to naturalistic stimuli (including experiences) entails building a model of the stimulus and searching for brain regions whose responses are consistent with the model.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have extended this approach with a clever twist.  Rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time.  By contrast, in our present work we used topic models and HMMs to construct an explicit stimulus model (i.e., the topic trajectory of the video).  When we searched for brain structures whose responses are consistent with the video's topic trajectory, we identified a network of structures that overlapped strongly with the ``long temporal receptive window'' network reported by the Hasson group~\citep[e.g., compare our Fig.~\ref{fig:brainz}A with the map of long temporal receptive window voxels in][]{LernEtal11}.  This provides support for the notion that part of the long temporal receptive window network may be maintaining an explicit model of the stimulus dynamics.  When we performed a similar analysis after swapping out the video's topic trajectory with the recall topic trajectories of each individual participant, this allowed us to identify brain regions whose responses (as the participants viewed the video) reflected how the video trajectory would be transformed in memory (as reflected by the recall topic trajectories).  The analysis revealed that the rMTL and vmPFC may play a role in this person-specific transformation from experience into memory.  The role of the MTL in episodic memory encoding has been well-reported~\citep[e.g., ][]{PallWagn02, DavaEtAl03, RangEtal04a, Dava06}.  Prior work has also implicated the medial prefrontal cortex in representing ``schema'' knowledge~\citep[i.e., general knowledge about the format of an ongoing experience given prior similar experiences; ][]{KestEtal12, SchlPres15, GilbMarl17, SpalEtal18}.  Integrating across our study and this prior work, one interpretation is that the person-specific transformations mediated (or represented) by the rMTL and vmPFC may reflect schema knowledge being leveraged, formed, or updated, incorporating ongoing experience into previously acquired knowledge.

When modeling memory experiments, often times events (or items) and their later memories are treated as binary (or categorical in the case of confidence ratings) and independent events. Our novel framework allows one to assess memory performance in a more continuous way (e.g. precision), as well as analyze the correlational structure of each encoding event to each memory event (e.g. distinctiveness). Further and importantly, it allows for consideration of the actual content of the experience/memories, which is not typically modeled. Leveraging this, using 2 novel memory metrics we find that the successful memory performance is related to 1) the \textit{precision} with which the participant recounts each event and 2) how \textit{distinctive} each recall event is (relative to the other recalled events). The first finding suggests to us that the accuracy of recall for \textit{any individual event} may predict the overall amount of information recovered by the participant.  The second finding suggests that remembering/describing events in a unique way (relative to other recalled events) is also related to the quantity of content recovered. Intriguingly, prior studies show that pattern separation, or the ability to discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging \citep{StarEtal10, YassEtal11c, YassStar11b}.  Future work might explore how/whether the novel metrics introduce here compare between cognitively impoverished groups and healthy controls.

% NOTE: do you know of a good reference for WAS and word2vec ?
While a large number of language models exist (e.g. WAS, LSA, word2vec, universal sentence encoder) \citep{LandEtal98, CerEtal18}, here we use topic models for a few reasons.  First, topic models capture the \textit{essence} of a text passage devoid of the specific set and order of words used.  This was an important feature of our model since different people may accurately recall a scene using very different language. Secondly, words can mean different things in different contexts (e.g. baseball bat vs. the animal bat), and topic models are robust to this since words can be part of multiple topics.  Lastly, topic models provide a straight forward to recover the weights for the particular words comprising a topic, allowing for easy interpretation of an event's contents (e.g. Fig.~\ref{fig:topics}). Other models such as Google's universal sentence encoder offer a context-sensitive encoding of text passages, but the encoding space is complex and non-linear and thus, recovering the original words used to fit the model is not straight forward. However, it's worth pointing out that our framework is divorced from the particular choice of language model. Moreover, many of the aspects of our framework could be swapped out for other choices. For example, the language model, the timeseries segmentation model and the video-recall matching function could all be customized for the particular problem. Indeed for some problems, recovery of the particular recall words may not be necessary, and thus other text-modeling approaches (such as universal sentence encoder) may be preferable. Future work will explore the influence of particular model choices on the framework's accuracy.

Our work has broad implications for how we characterize and assess memory in real-world settings such as the classroom or physician's office.  For example, the most commonly used classroom evaluation tools involve computing the proportion of correctly answered exam questions.  Our work indicates that this approach is only loosely related to what educators might really want to measure: how well did the students understand the key ideas presented in the course?  One could apply the computational framework we developed to construct topic trajectories for the video and participants' recalls to build explicit content models of the course material and exam questions.  This approach would provide a more nuanced and specific view into which aspects of the material students had learned well (or poorly).  In clinical settings, memory measures that incorporate such explicit content models might also provide more direct evaluations of patients' memories.

>>>>

The above results indicate that both the structure of the original video and participants' recalls of the video exhibit event boundaries that can be identified automatically by characterizing the dynamic content using a shared topic model and segmenting the content into events using HMMs.  Next, we asked whether some correspondence might be made between the specific content of the events the participants experienced in the video, and the events they later recalled.  One approach to linking the experienced (video) and recalled events is to label each recalled event as matching the video event with the most similar (i.e., most highly correlated) topic vector (Figs.~\ref{fig:model}G, \matchmats).  This yields a sequence of ``presented'' events from the original video, and a sequence of (potentially differently ordered) ``recalled'' events for each participant.  Analogous to classic list-learning studies, we can then examine participants' recall sequences by asking which events they tended to recall first~\citep[probability of first recall; Fig.~\ref{fig:list-learning}A;][]{WelcBurn24, PostPhil65, AtkiShif68}; how participants most often transition between recalls of the events as a function of the temporal distance between them~\citep[lag-conditional response probability; Fig.~\ref{fig:list-learning}B;][]{Kaha96}; and which events they were likely to remember overall~\citep[serial position recall analyses; Fig.~\ref{fig:list-learning}C;][]{Murd62a}. Interestingly, for two of these analyses (probability of first recall and lag-conditional response probability curves) we observe patterns comparable to classic effects from the list-learning literature: namely, a higher probability of initiating recall with the first event in the sequence (Fig.~\ref{fig:list-learning}A) and a higher probability of transitioning to neighboring events with an asymmetric forward bias (Fig.~\ref{fig:list-learning}C). In contrast, we do not observe a pattern comparable to the serial position effect (Fig.~\ref{fig:list-learning}C), but rather we see higher memory for specific events distributed somewhat evenly throughout the video.

@@

Our work casts remembering as reproducing (behaviorally and neurally) the topic trajectory, or shape, of an experience.  This view draws inspiration from prior work aimed at elucidating the neural and behavioral underpinnings of how we process dynamic naturalistic experiences and remember them later.  One approach to identifying neural responses to naturalistic stimuli (including experiences) entails building a model of the stimulus and searching for brain regions whose responses are consistent with the model.  In prior work, a series of studies from Uri Hasson's group~\citep{LernEtal11, SimoEtal16, ChenEtal17, BaldEtal17, ZadbEtal17} have extended this approach with a clever twist: rather than building an explicit stimulus model, these studies instead search for brain responses (while experiencing the stimulus) that are reliably similar across individuals.  So called \textit{inter-subject correlation} (ISC) and \textit{inter-subject functional connectivity} (ISFC) analyses effectively treat other people's brain responses to the stimulus as a ``model'' of how its features change over time.  By contrast, in our present work we used topic models and HMMs to construct an explicit stimulus model (i.e., the topic trajectory of the video).  When we searched for brain structures whose responses are consistent with the video's topic trajectory, we identified a network of structures that overlapped strongly with the ``long temporal receptive window'' network reported by the Hasson group~\citep[e.g., compare our Fig.~\ref{fig:brainz}A with the map of long temporal receptive window voxels in][]{LernEtal11}.  This provides support for the notion that part of the long temporal receptive window network may be maintaining an explicit model of the stimulus dynamics.  When we performed a similar analysis after swapping out the video's topic trajectory with the recall topic trajectories of each individual participant, this allowed us to identify brain regions whose responses (as the participants viewed the video) reflected how the video trajectory would be transformed in memory (as reflected by the recall topic trajectories).  The analysis revealed that the rMTL and vmPFC may play a role in this person-specific transformation from experience into memory.  The role of the MTL in episodic memory encoding has been well-reported~\citep[e.g., ][]{PallWagn02, DavaEtAl03, RangEtal04a, Dava06, WiltSilv07, DianEtal07, KestEtal13}.  Prior work has also implicated the medial prefrontal cortex in representing ``schema'' knowledge~\citep[i.e., general knowledge about the format of an ongoing experience given prior similar experiences; ][]{KestEtal12, KestEtal13, SchlPres15, GilbMarl17, SpalEtal18}.  Integrating across our study and this prior work, one interpretation is that the person-specific transformations mediated (or represented) by the rMTL and vmPFC may reflect schema knowledge being leveraged, formed, or updated, incorporating ongoing experience into previously acquired knowledge.
% Here's a bunch of citations to possibly add to this last sentence, though not sure if we should because the sentence is framed as a "new idea" to come out of this study:
% Takashima et al., 2007 -- vmPFC involved in stabilizing/reorganizing retrieval networks
% Tse et al., 2007; Zeithamova and Preston, 2010; Tse et al., 2011 -- vmPFC involved in integrating featurally similar experiences
% Nieuwenhuis and Takashima, 2011 -- vmPFC activity linked to evaluating memories' relevance to current situation
% Kuhl et al., 2007; Wimber et al., 2008, 2009, 2011, 2015 -- vmPFC and vlPFC inhibit contextually irrelevant memories during retrieval
% -Paxton

In extending classical free recall analyses to our naturalistic memory framework, we recovered two patterns of recall dynamics similar to those central to list-learning studies: a high probability of initiating recall with the first video event (Fig.~\ref{fig:list-learning}A) and a strong bias toward transitioning from recalling a given event to recalling the event immediately following it (Fig.~\ref{fig:list-learning}B).  However, equally noteworthy are the typical free recall results not recovered in these analyses, as each highlights a fundamental difference between list-learning studies and naturalistic memory paradigms like that employed in the present study.  The most noticeable departure from hallmark free recall dynamics in these findings is the apparent lack of a serial position effect in Figure \ref{fig:list-learning}C, which instead shows greater and lesser recall probabilities for events distributed across the video.  Free recall experiments' stimuli most often comprise lists of simple, common words, presented to participants in a random order. \citep[In fact, numerous word pools have been developed for precisely this purpose; e.g.,][]{FrieEtal82}.  These stimulus qualities enable two assumptions that are central to word list analyses, but frequently do not hold for real-world experiences.  First, researchers conducting free recall studies may assume that the content at each presentation index is essentially equal, and does not bear qualities that would affect participants' abilities to remember it above that of others.  Such is rarely the case with real-world experiences or experiments meant to approximate them, and the effects of both intrinsic and observer-dependent factors on memorability are well-established \citep[for review see][]{ChunTurk07, ByliEtal15, TyngEtal17}.  Second, the random ordering of list items ensures that (across participants, on average) there is no relationship between the thematic similarity of stimuli and their presentation positions---in other words, two semantically related words are no more likely to be presented next to each other than at opposite ends of the list.  In most cases, the exact opposite is true of real-world episodes.  Our internal thoughts, our actions, and the physical state of the world around us all tend to follow a direct, causal progression.  And as a result, each moment of our experience tends to be inherently more similar to surrounding moments than to those in the distant past or future.  Memory literature has termed this strong temporal autocorrelation ``context," and in various media that depict real-world events (e.g., movies and written stories), we recognize it as a \textit{narrative structure}.  While random word lists (by definition) have no such structure, the logical progression between ideas and actions in a naturalistic scenario prompts the rememberer to recount presented events in order, from the beginning.  This pattern is reflected in our findings' second departure from typical free recall dynamics: a lack of increased probability of first recall for end-of-sequence events (Fig.~\ref{fig:list-learning}A).

When modeling memory experiments, often times events (or items) and their later memories are treated as binary (or categorical in the case of confidence ratings) and independent events. Our novel framework allows one to assess memory performance in a more continuous way (e.g. precision), as well as analyze the correlational structure of each encoding event to each memory event (e.g. distinctiveness). Further and importantly, it allows for consideration of the actual content of the experience/memories, which is not typically modeled. Leveraging this, using two novel memory metrics we find that the successful memory performance is related to 1) the \textit{precision} with which the participant recounts each event and 2) the \textit{distinctiveness} of each recall event (relative to the other recalled events). The first finding suggests that the information retained for \textit{any individual event} may predict the overall amount of information retained by the participant.  The second finding suggests that the ability to distinguish between temporally or semantically similar content is also related to the quantity of content recovered. Intriguingly, prior studies show that pattern separation, or the ability to discriminate between similar experiences, is impaired in many cognitive disorders as well as natural aging \citep{StarEtal10, YassEtal11c, YassStar11b}.  Future work might explore how/whether the novel metrics introduce here compare between cognitively impoverished groups and healthy controls.

% NOTE: do you know of a good reference for WAS and word2vec ?
% SteyEtal04 = WAS; MikoEtal13 = word2vec
% -Paxton
While a large number of language models exist \citep[e.g., WAS, LSA, word2vec, universal sentence encoder;][]{SteyEtal04, LandEtal98, MikoEtal13, CerEtal18}, here we use latent dirichlet allocation (LDA)-based topic models for a few reasons.  First, topic models capture the \textit{essence} of a text passage devoid of the specific set and order of words used.  This was an important feature of our model since different people may accurately recall a scene using very different language. Second, words can mean different things in different contexts (e.g. ``bat" as the act of hitting a baseball, the object used for that action, or a flying mammal).  Topic models are robust to this, allowing words to exist as part of multiple topics.  Last, topic models provide a straightforward means to recover the weights for the particular words comprising a topic, enabling easy interpretation of an event's contents (e.g. Fig.~\ref{fig:topics}). Other models such as Google's universal sentence encoder offer a context-sensitive encoding of text passages, but the encoding space is complex and non-linear, and thus recovering the original words used to fit the model is not straightforward. However, it's worth pointing out that our framework is divorced from the particular choice of language model. Moreover, many of the aspects of our framework could be swapped out for other choices. For example, the language model, the timeseries segmentation model and the video-recall matching function could all be customized for the particular problem. Indeed for some problems, recovery of the particular recall words may not be necessary, and thus other text-modeling approaches (such as universal sentence encoder) may be preferable. Future work will explore the influence of particular model choices on the framework's accuracy.

Our work has broad implications for how we characterize and assess memory in real-world settings, such as the classroom or physician's office.  For example, the most commonly used classroom evaluation tools involve simply computing the proportion of correctly answered exam questions.  Our work indicates that this approach is only loosely related to what educators might really want to measure: how well did the students understand the key ideas presented in the course?  Under this typical framework of assessment, the same exam score of 50\% could be ascribed to two very different students: one who attended the full course but struggled to learn more than a broad overview of the material, and one who attended only half of the course but understood the material perfectly.  Instead, one could apply our computational framework to build explicit content models of the course material and exam questions.  This approach would provide a more nuanced and specific view into which aspects of the material students had learned well (or poorly).  In clinical settings, memory measures that incorporate such explicit content models might also provide more direct evaluations of patients' memories.
